---
title: "To what extent does mental health alerts, length of segregation placement, gender and race group predict the suicide risk alert in Ontario correctional facilities in 2018"
author: "Alex_nguyen"
date: "2026-03-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
seg = read.csv("solgen-individuals-in-seg-2018.csv")
seg = seg %>% rename(Seg_length = Segregation.placement.length..in.days.) %>% 
  rename(Reason = Primary.reason.for.segregation)
```

# Introduction

This data set contains individual-level records of incarcerated persons placed in segregation within Ontario correctional facilities. Each row represents one individual and includes demographic characteristics such as gender, age category, race group, and faith group. The data also captures clinical indicators including mental health alerts, suicide risk alerts, and suicide watch status. The key outcome variables measure segregation placement length in days and the primary reason for segregation placement.

## Segregation and Suicide Risk in Ontario

Segregation involves confining imprisoned individuals to their cells for 20-24 hours per day, with limited human contact. In Ontario, the use of segregation has been a subject of significant policy debate due to concerns about its psychological impacts, especially on vulnerable populations.

Research has shown that individuals with mental health conditions are at elevated risk of suicide, and that segregation placement may exacerbate this risk. However, questions remain about the extent to which mental health status, segregation length, gender, and race independently predict suicide risk alerts in correctional settings.

This study addresses the following research question: To what extent do mental health alerts, length of segregation placement, gender, and race group predict suicide risk alerts in Ontario correctional facilities in 2018?

### Relationship between variables

Before conducting formal hypothesis testing and logistic regression, we first examine the bivariate relationships between each predictor and the outcome variable, suicide risk alerts. These visualizations help assess whether relationships align with expectations, and inform which variables is important in the regression model. We begin by examining the relationship between mental health alerts and suicide risk alerts.

#### Mental Health Alerts and Suicide Risk

```{r}
mental_suicide = seg %>% group_by(Mental.health.alert, Suicide.risk.alert) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Mental.health.alert) %>% mutate(percentage = count/sum(count) * 100)

ggplot(mental_suicide, aes(x = factor(Mental.health.alert), 
                           y = percentage, 
                           fill = factor(Suicide.risk.alert, labels = c("No Suicide Risk", "Suicide Risk")))) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Figure 1. Proportion of Suicide Risk Alerts by Mental Health Status",
       x = "Mental Health Alert", 
       y = "Proportion",
       fill = "Suicide Risk") +
  scale_y_continuous(labels = scales::percent) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Figure 1 displays the proportion of individuals with suicide risk alerts stratified by mental health status. The stacked bar chart allows for direct comparison between two groups, including individuals without mental health alerts and individuals with mental health alerts.

Among individuals without mental health alerts, approximately 25% have a suicide risk alert, meaning that three-quarters of this group show no suicide risk flag. On the other hand, among individuals with mental health alerts, approximately 50% have a suicide risk alert. We can observe a pattern which reveals a strong positive association between mental health status and suicide risk identification. Individuals flagged for mental health concerns are more likely to also be flagged for suicide risk compared to those without mental health alerts. The difference of approximately 25 percentage suggests that mental health status may be the most important predictors of suicide risk in this sample.
