---
title: "To what extent does mental health alerts, length of segregation placement, gender and race group predict the suicide risk alert in Ontario correctional facilities in 2018"
author: "Alex_nguyen"
date: "2026-03-22"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, include=FALSE}
library(rattle)
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(knitr)
set.seed(67)
seg = read.csv("solgen-individuals-in-seg-2018.csv")
seg = seg %>% rename(Seg_length = Segregation.placement.length..in.days.) %>% 
  rename(Reason = Primary.reason.for.segregation)
```

# Introduction

This data set contains individual-level records of incarcerated persons placed in segregation within Ontario correctional facilities. Each row represents one individual and includes demographic characteristics such as gender, age category, race group, and faith group. The data also captures clinical indicators including mental health alerts, suicide risk alerts, and suicide watch status. The key outcome variables measure segregation placement length in days and the primary reason for segregation placement.

## Segregation and Suicide Risk in Ontario

Segregation involves confining imprisoned individuals to their cells for 20-24 hours per day, with limited human contact. In Ontario, the use of segregation has been a subject of significant policy debate due to concerns about its psychological impacts, especially on vulnerable populations.

Research has shown that individuals with mental health conditions are at elevated risk of suicide, and that segregation placement may exacerbate this risk. However, questions remain about the extent to which mental health status, segregation length, gender, and race independently predict suicide risk alerts in correctional settings.

This study addresses the following research question: To what extent do mental health alerts, length of segregation placement, gender, and race group predict suicide risk alerts in Ontario correctional facilities in 2018?

### Relationship between variables

Before conducting formal hypothesis testing and logistic regression, we first examine the bivariate relationships between each predictor and the outcome variable, suicide risk alerts. These visualizations help assess whether relationships align with expectations, and inform which variables is important in the regression model. We begin by examining the relationship between mental health alerts and suicide risk alerts.

#### Mental Health Alerts and Suicide Risk

```{r echo=FALSE, fig.align="center", fig.width=5, fig.height=3}
mental_suicide = seg %>% group_by(Mental.health.alert, Suicide.risk.alert) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Mental.health.alert) %>% mutate(percentage = count/sum(count) * 100)

ggplot(mental_suicide, aes(x = factor(Mental.health.alert), 
                           y = percentage, 
                           fill = factor(Suicide.risk.alert, labels = c("No Suicide Risk", "Suicide Risk")))) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Figure 1. Proportion of Suicide Risk Alerts by Mental Health Status",
       x = "Mental Health Alert", 
       y = "Proportion",
       fill = "Suicide Risk") +
  scale_y_continuous(labels = scales::percent) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = rel(0.8)))
```

Figure 1 displays the proportion of individuals with suicide risk alerts stratified by mental health status. The stacked bar chart allows for direct comparison between two groups, including individuals without mental health alerts and individuals with mental health alerts.

Among individuals without mental health alerts, approximately 25% have a suicide risk alert, meaning that three-quarters of this group show no suicide risk flag. On the other hand, among individuals with mental health alerts, approximately 50% have a suicide risk alert. We can observe a pattern which reveals a strong positive association between mental health status and suicide risk identification. Individuals flagged for mental health concerns are more likely to also be flagged for suicide risk compared to those without mental health alerts. The difference of approximately 25 percentage suggests that mental health status may be the most important predictors of suicide risk in this sample.

This align with what is expected. As mental health conditions can be a risk factors for suicidal behavior. In the given settings, where mental health can be negated, it can easily lead to high suicidal risk.

#### Segregation length and Suicide Risk \newline

Before examining the relationship between segregation length and suicide risk, we must first address the presence of extreme outliers in the segregation length variable. Table 1 summarizes the outlier statistics for individuals with and without suicide risk alerts.

```{r echo=FALSE}
getting_outliers = function(seg_length){
  Q1 = quantile(seg_length, 0.25, na.rm = TRUE)
  Q3 = quantile(seg_length, 0.75, na.rm = TRUE)
  IQR = Q3 - Q1
  lower_bound = Q1 - 1.5 * IQR
  upper_bound = Q3 + 1.5 * IQR
  outliers = seg_length[seg_length > upper_bound | seg_length < lower_bound]
  return(list(outliers = outliers, upper_bound = upper_bound))
  
}

seg_no_risk = seg$Seg_length[seg$Suicide.risk.alert == "No"]
seg_yes_risk = seg$Seg_length[seg$Suicide.risk.alert == "Yes"]

outlier_no = getting_outliers(seg_no_risk)
outlier_yes = getting_outliers(seg_yes_risk)


outlier_table = data.frame(
  Suicide_Risk = c("No", "Yes"),
  Total_N = c(length(seg_no_risk), length(seg_yes_risk)),
  Outlier_Count = c(length(outlier_no$outliers), length(outlier_yes$outliers)),
  Outlier_Percent = c(round(length(outlier_no$outliers) / length(seg_no_risk) * 100, 1),
                      round(length(outlier_yes$outliers) / length(seg_yes_risk) * 100, 1)),
  Outlier_Range = c(paste(round(min(outlier_no$outliers), 1), "-", round(max(outlier_no$outliers), 1)),
                    paste(round(min(outlier_yes$outliers), 1), "-", round(max(outlier_yes$outliers), 1)))
)
kable(outlier_table, 
      caption = "Outlier Summary by Suicide Risk Status",
      col.names = c("Suicide Risk", "Total N", "Outlier Count", "Outlier (%)", "Outlier Range"),
      align = c('l', 'c', 'c', 'c', 'c'))
```

Table 1 reveals that outliers are present in both groups at very high rates such as 14.0% of individuals without suicide risk alerts and 14.4% of individuals with suicide risk alerts are classified as outliers. This indicates that extreme segregation lengths are not disproportionately concentrated in either group. The outlier ranges extend up to approximately 600 days in both categories, demonstrating that while some individuals experience exceptionally long segregation placements, these extreme cases occur with similar frequency regardless of suicide risk status.

As documented in Table 1, outliers are prevalent in both groups (14.0% and 14.4%), necessitating a truncated view of the data to focus on the central tendency. Figure 4 presents the distribution of segregation placement length by suicide risk status, with the y-axis limited to the 85th percentile to reduce visual distortion caused by extreme outliers.

```{r echo=FALSE, fig.align="center", fig.width=5, fig.height=3}
ggplot(seg, aes(x = factor(Suicide.risk.alert, labels = c("No", "Yes")), 
                          y = Seg_length)) +
  geom_boxplot(fill = c("lightblue", "salmon")) +
  coord_cartesian(ylim = c(0, quantile(seg$Seg_length, 0.85))) +
  labs(title = "Figure 2. Segregation Placement Length by Suicide Risk Status",
       x = "Suicide Risk Alert", 
       y = "Days in Segregation") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = rel(0.8)))
```

The boxplot reveals that individuals with suicide risk alerts have a mean segregation length only approximately 2 days higher than those without. Given the small magnitude of this difference and the substantial overlap in the interquartile ranges between groups, segregation length does not appear to be a strong or meaningful predictor of suicide risk status in this population. This suggests that other factors, such as mental health alerts, likely play a more important role in identifying individuals at risk of suicide later on when we do the regression.

#### **Hypothesis testing**

Having explored the relationships between each predictor and suicide risk alerts through visualizations, we now turn into doing hypothesis testing between different groups. Hypothesis testing allows us to determine whether observed differences between groups are significant or likely due to random chance.

##### Permutation Test for Gender Differences

This section focuses on the relationship between gender and suicide risk alerts. Specifically, we test whether the proportion of individuals with suicide risk alerts differs between males and females in Ontario correctional facilities in 2018. Because the data may not meet the normality assumptions required for a standard two-sample proportion test, we employ a permutation test. This approach involves randomly shuffling the gender labels across observations many times (10,000 permutations) to generate a distribution of differences expected under the null hypothesis. The observed difference is then compared to this null distribution to determine the significance.

###### Null and Alternative Hypotheses

$$H_0: p_m = p_f$$

$$H_1: p_m \neq p_f$$

Where $p_m$ represents the true proportion of males with suicide risk alerts, and $p_f$ represents the true proportion of females with suicide risk alerts.

```{r echo=F, fig.align="center", fig.width=5, fig.height=3}
seg_perm_test = seg %>% mutate(Suicide_binary = case_when(Suicide.risk.alert == 'Yes' ~1, TRUE ~ 0)) %>% select(Suicide_binary, Gender)
obs.diff = mean(seg_perm_test$Suicide_binary[seg_perm_test$Gender== "Male"]) - mean(seg_perm_test$Suicide_binary[seg_perm_test$Gender== "Female"])

perm.function = function(){
  temp_group_ind = sample(seg_perm_test$Gender)
  d = mean(seg_perm_test$Suicide_binary[temp_group_ind== "Male"]) - mean(seg_perm_test$Suicide_binary[temp_group_ind== "Female"])
  return(d)
}

perm_diff = replicate(5000, perm.function())
ci = quantile(perm_diff, c(0.025, 0.975))
perm_df = data.frame(diff = perm_diff)
ggplot(perm_df, aes(x = diff)) +
  geom_histogram(fill = "lightgray", color = "white", bins = 30) +
  geom_vline(xintercept = obs.diff, color = "blue", linewidth = 1.2) +
  geom_vline(xintercept = ci, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "Figure 3: Permutation Distribution of Gender Differences in Suicide Risk",
       x = "Difference in Proportion (Male - Female)",
       y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),     
    axis.title = element_text(size = 9),  
    axis.text = element_text(size = 8),        
    legend.text = element_text(size = 8)       
  )
#mean(abs(perm_diff) > abs(obs.diff)) #p-value
#obs.diff

##seg_perm_male = seg_perm_test %>% filter(Gender == "Male")
##seg_perm_female = seg_perm_test %>% filter(Gender == "Female")
##mean(seg_perm_male$Suicide_binary)
##mean(seg_perm_female$Suicide_binary)
##mean_diff = mean(seg_perm_male$Suicide_binary) - mean(seg_perm_female$Suicide_binary)
##mean_diff
# These numbers are the current mean. 
```

The permutation test yielded the following results:

```{r echo=FALSE}
results_table <- data.frame(
  Statistic = c("Observed Difference (Male - Female)", 
                "95% Confidence Interval", 
                "P-value"),
  Value = c("-0.053", "(-0.053, 0.051)", "0.045")
)
kable(results_table, caption = "Table 2: Permutation Test Results for Gender Differences in Suicide Risk")
```

Figure 3 displays the permutation distribution of differences in suicide risk proportions between males and females under the null hypothesis. The blue vertical line represents the observed difference in the actual data, while the red dashed lines indicate the 95% confidence interval.

The 95% confidence interval for the difference in proportions ranges from -0.053 to 0.051. This interval crosses zero, indicating that the true population difference could be negative, zero, or positive. However, the interval is narrow and nearly centered around zero, suggesting that if a true difference exists, it is likely small in magnitude.

The p-value of 0.0442 is less than the significance level of 0.05. This means that if the null hypothesis were true (no difference between genders), the probability of observing a difference as extreme as the one we observed is approximately 4.4%.

Based on the p-value of 0.0442, we reject the null hypothesis at the α = 0.05 significance level. There is statistically significant evidence that the proportion of suicide risk alerts differs between males and females in Ontario correctional facilities in 2018.

However, several important caveats warrant discussion. First, while the result is statistically significant, the 95% confidence interval includes zero, indicating uncertainty about the direction of the effect. This apparent contradiction occurs because the p-value is very close to the significance threshold, and the confidence interval is calculated using a different method. Second, the magnitude of any true gender difference appears to be small, as the confidence interval spans only approximately 5 percentage points in either direction.

From a practical perspective, while gender may be a statistically significant predictor of suicide risk alerts, its substantive importance may be limited. The small effect size suggests that gender alone is unlikely to be a strong or reliable predictor of suicide risk in correctional settings. Other factors, such as mental health status, likely play a more meaningful role in identifying individuals at risk of suicide.

```{r}

seg_regression = seg %>% 
  mutate(Suicide_binary = ifelse(Suicide.risk.alert == "Yes",1,0),
         Mental_binary = ifelse(Mental.health.alert == "Yes",1,0),
         Gender_binary = ifelse(Gender == "Male",1,0))
m = glm(Suicide_binary ~ Gender_binary+Mental_binary + Seg_length + Reason, family = binomial, data = seg_regression)
summary(m)
library(pROC)
p = predict(m, type = "response")
roc_logit = roc(seg_regression$Suicide_binary ~ p)
ggroc(roc_logit, color="red", size=2)
auc(roc_logit)
```

```{r}
library(rpart)
m2 = rpart(Suicide_binary ~ Gender_binary + Mental_binary + Seg_length +Reason,
data = seg_regression, method="class")
library(rattle)
fancyRpartPlot(m2)
```
